Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-06-05 paper/paper.tex

We have shown that it is possible to obtain photometric parallaxes for
distant stars in the \gaia\ \tgas\ data without any use of physical stellar
models, nor models of the Milky Way.
We used the geometric parallaxes to calibrate a photometric
model that is purely statistical, which is a model of the data rather than
a model of stars \foreign{per se}.
This opens up the possibility of completing the goals of the \gaia\ Mission
without building in unnecessary assumptions about the mechanical properties
of stars or the Galaxy.

@@

Although we have not performed principled hierarchical Bayesian inference in this
project, we built our prior for each star's parallax by
performing a statistically responsible deconvolution
of the \gaia\ \tgas\ data that is justifiable under a clear set of
assumptions.
This deconvolution, using a reasonable noise model and an
assumption of stationarity, shrinks the parallax uncertainties, most dramatically for the stars measured at lower signal-to-noise.
Because the method we use, \xd, which is
an Empirical Bayesian maximum-marginalized-likelihood estimator, accounts for
heteroskedastic noise, we were able to build our photometric parallax
model---which is a model of the \cmd---using
all the data, not just the data with the highest signal-to-noise.
So, the model for the \cmd\ we built is representative for
our subset of the \tgas\ Catalog.
Because our model is representitive of our full subset catalog, we expect any (sensible) parallax estimates
we generate from our parallax posterior pdfs to be only weakly biased.

\subsection{Comparison with other priors}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{comparePrior.png}
\caption{Comparison with common prior: A visualization of the posterior pdf for a subsection of TGAS stars, shown as the expectation value and the sqrt of the variance as the uncertainty. On the left is the posterior pdf using our \cmd\ prior, and on the right is the posterior pdf when using the exponentially declining spatial density, a common prior used when inferring parallaxes. The scale of the exponentially declining spatial density prior is 1.35 kpc, the optimal value found in \cite{astraatmadja16a}. Using the \cmd\ as a prior generates a posterior pdf that has smaller variance.}
\label{fig:comparePrior}
\end{figure}

There are other options for prior information to include when inferring distances to stars.
One option is the exponentially declining spatial density (EDSD) of stars \citep{astraatmadja16a}.
It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so the posteriors are also very smooth.
It is a fairly weak prior.
It sets a smooth scale distance beyond which the posterior struggles to maintain much probability, and also converts a negative parallax likelihood into a posterior of very small parallax.
A comparison of the posteriors using the two different priors, our \cmd\ prior and the EDSD, is shown in \figurename~\ref{fig:comparePrior}.
Here the parallax posterior pdf expectation value and square root of the variance is converted into an absolute magnitude and visualized as the \cmd.
The left panel visualizes results using our \cmd\ prior, and the right panel visualizes results using the EDSD prior.
The variance for the \cmd\ prior is significantly smaller than the variance for the EDSD prior.
In general, it is a fairly broad prior so the median \gaia\ star has a posterior variance equal to the likelihood variance.
It has the largest effect on the most noisy and negative parallaxes, and has the nice property of bringing these noisy or negative measurements to reasonable parallaxes.

\subsection{What's That Feature?}
The denoised expectation values of the parallax projected into the \cmd\ show interest features.
We visualize this \cmd\ again in \figurename~\ref{fig:wtf} pointing out some interesting features.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{whatsThatFeature.pdf}
\caption{Features in the \cmd: The expectation values from the posterior pdfs of the true parallax}
\label{fig:wtf}
\end{figure}

The lower main sequence is very narrow while the upper main sequence
is much wider. This may be an age variation on the main sequence mixed
with a malmquist bias of seeing a much larger volume of upper main
sequence stars. There is a plausible binary sequence of stars with
about twice the brightness as the main sequence. The helium burning
red clump is prominent for the volume DR1 probed, and the red giant
branch is fairly straight and narrow. The main sequence turn off is
noticeable and also surprisingly narrow, possibly a reflection of the
star formation history of the local volume also mixed with a malmquist
bias.

@@

  relatively lower signal to noise objects compared with the main

@@

  detail, because the low and high signal to noise objects have to be

@@

  the \cmd\ from the data is that we made no use of any kind of
  \tgas\ selection function or completeness estimate or

@@

  be easily optimized. We performed some experiments with

>>>>

We have shown that it is possible to obtain photometric parallaxes for
distant stars in the \gaia\ \tgas\ catalog without any use of physical stellar
models, nor stellar density models of the Milky Way.
We used the geometric parallaxes to calibrate a photometric
model that is purely statistical, which is a model of the data rather than
a model of stars \foreign{per se}.
This opens up the possibility of completing the goals of the \gaia\ Mission
without building in unnecessary assumptions about the mechanical properties
of stars or the Galaxy.

@@

Although we have not performed principled hierarchical Bayesian inference in this
project, we built our prior for each star's parallax by
performing a statistically responsible deconvolution
of the \gaia\ \tgas\ data that is justifiable under a clear set of
assumptions.
This deconvolution, using a reasonable noise model and an
assumption of stationarity, shrinks the parallax uncertainties, most dramatically for the stars measured at lower signal-to-noise.
Because the method we use, \xd, which is
an Empirical Bayesian maximum-marginalized-likelihood estimator, accounts for
heteroskedastic noise, we are able to build our photometric parallax
model---which is a model of the \cmd---using
all the data, not just the data with the highest signal-to-noise.
So, the generated model for the \cmd\ is representative for
our subset of the \tgas\ Catalog.
Because our model is representitive of our full subset catalog, we expect any (sensible) parallax estimates
we generate from our parallax posterior pdfs to be only weakly biased.

\subsection{Comparison with other priors}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{comparePrior.png}
\caption{Comparison with a common spatial prior: A visualization of the posterior pdf for a subsection of \tgas\ stars, shown as the expectation value and the sqrt of the variance as the uncertainty. Left panel shows the posterior pdf using our \cmd\ prior, right panel shows the posterior pdfs when using the exponentially declining spatial density, a common space-density prior used when inferring distances. The scale of the exponentially declining spatial density prior is $1.35~{\rm kpc}$, the optimal value found in \cite{astraatmadja16a}. Using the \cmd\ as a prior generates a posterior pdf that has smaller variance.}
\label{fig:comparePrior}
\end{figure}

There are other options for prior information to include when inferring distances to stars.
One option is the exponentially declining spatial density (EDSD) of stars \citep{astraatmadja16a}.
It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so that the posteriors are also very smooth.
It is a fairly weak prior.
It sets a smooth scale distance beyond which the posterior struggles to maintain much probability, and also converts a negative parallax likelihood into a posterior of very small parallax.
\figurename~\ref{fig:comparePrior} shows a comparison of the EDSD prior with our \cmd\ prior.
Here the parallax posterior pdf expectation value and square root of the variance is converted into an absolute magnitude and visualized as the \cmd.
The left panel visualizes results using our \cmd\ prior, and the right panel visualizes results using the EDSD prior.
The variance for the \cmd\ prior is significantly smaller than the variance for the EDSD prior.
In general, it is a fairly broad prior so the median \gaia\ star has a posterior variance equal to the likelihood variance.
It has the largest effect on the most noisy and negative parallaxes, and has the nice property of bringing these noisy or negative measurements to reasonable parallaxes.

\subsection{What's That Feature?}
The denoised expectation values of the parallax projected into the \cmd\ show many interesting and familiar features; we visualize this \cmd\ again in \figurename~\ref{fig:wtf} and highlight some of these features.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{whatsThatFeature.pdf}
\caption{Features in the \cmd: The expectation values from the posterior pdfs of the true parallax}
\label{fig:wtf}
\end{figure}

The lower main sequence is very narrow while the upper main sequence
is much wider. This may be an age variation on the main sequence mixed
with a Malmquist bias of seeing a much larger volume of upper main
sequence stars. There is a plausible binary sequence of stars with
about twice the brightness as the main sequence. The helium burning
red clump is prominent for the volume DR1 probed, and the red giant
branch is fairly straight and narrow. The main sequence turn off is
noticeable and also surprisingly narrow, possibly a reflection of the
star formation history of the local volume also mixed with a Malmquist
bias.

@@

  relatively lower signal-to-noise objects compared with the main

@@

  detail, because the low and high signal-to-noise objects have to be

@@

  the \cmd\ from the data is that we have not used a
  \tgas\ selection function, completeness estimate, or

@@

  be easily optimized. We performed some experiments
